% Abstract (does not appear in the Table of Contents)
\chapter*{Abstract}%

The traditional geometry of classical mechanics is symplectic geometry. Autonomous mechanical systems defined on symplectic manifolds do not support energy dissipation. In this thesis, we investigate the correct differential geometric structures that support dissipation in mechanical and the analogous economic engineering systems. Furthermore, we propose an alternative representation for two-dimensional linear dynamical systems in the form of split-quaternions.

First, we investigate the role of contact Hamiltonian systems for the description of dissipative mechanical systems. We use thermodynamic theory to derive the contact structure, and modify the existing approaches to this problem to assign them with a direct physical interpretation, suitable for engineering applications. The damped harmonic oscillator with one or two dampers is used as an exemplary mechanical system. We also investigate the symplectification of the associated manifold to cast this system on a symplectic manifold with an additional Liouville structure. We show that the symplectified Hamiltonian system is directly equivalent to the well-known Caldirola-Kanai model for the damped oscillator.

Second, we show that contact manifolds fall short when applied to more general, multi-degree of freedom mechanical systems because of they derive the symplectic isomorphism from the contact 1-form. A modified geometric structure, an instance of the more general class of Jacobi structures, is then used to establish a framework that captures any mechanical system.  

Thirdly, we propose an alternative for the traditional matrix representation of two-dimensional linear mechanical systems. We show that the natural properties of split-quaternions translate directly to the qualitative features of their assocatiated dynamical systems and therefore conveniently facilitate the classification of fixed points. The system solution is also readily obtained through the split-quaternion representation. Additionally, we again employ the harmonic oscillator with two dampers as an exemplary system to relate the geometry of the solution trajectories directly to the split-quaternion representation of the system. In particular, we investigate the geometry of underdamped mechanical systems in the Lorentzian 3-space using models of the hyperbolic plane.

 
