\chapter{Preface}
\thispagestyle{empty}
This thesis stands at the end of a long and twisty road with many dead ends. Indeed, even four months ago, I could not have envisioned what the final result turned out to be. This learned me an important lesson: that good results often come unexpected, but they never come for free. It takes some courage to invest so much time in studying subjects that are completely unfamiliar, and seem daunting at first. And even then, it takes innumerable mistakes to finally arrive at the correct result. 
However, this makes me all the more proud to present my work in this thesis; I sincerely hope that the reader shares in some of my enthusiasm, and finds it at least partially interesting.

Of course, I would never have succeeded in this task were it not for the support and assistance that I received along the way. First of all, I want to express my sincere gratitude to my supervisor prof. em. Mendel. The word `supervisor' definitely does not do justice to the substantial contributions that prof. em. Mendel made to this thesis, for most of the ideas presented are at least partially his, or the result of our frequent discussions. His deep intuition often directed us to the right answer, even if the intermediate steps were completely unknown to either of us. I can only be deeply grateful for all his time, inspiring ideas, and unwavering enthusiasm over the past eighteen months. 

Second, I want to thank my parents for their intensive support, not only during my thesis but during my entire study, and for allowing me to pursue my dreams and ambitions. I have come to appreciate it deeply during the past year.

Third, I want to express how happy I am to have spent the past seven years in Delft with such excellent people. In particular, I want to thank my housemates Wout, Tim, Bram, Fréderic, Tine and Evert for making living in Delft so enjoyable --- may the Bronx flourish until eternity. Furthermore, my gratitude goes out to my fellow NunaX teammates and the the people of my student society, with whom I have built invaluable friendships over time.

Finally, I thank my peers in the economic engineering group for all their valuable feedback and suggestions, and for enduring the theoretical drivel they had to sit through during my presentations.

Mercikes!

\begin{flushright}
    Emiel Legrand
\end{flushright}
Delft, \mscdate



