\chapter{Split-Quaternion Representations of Dynamical Systems}
\label{chap:quaternion}
In this chapter, we propose split-quaternions as an alternative representation of linear two-dimensional dynamical systems instead of the traditional state-space representation. This representation is based on the fact that the algebras of split-quaternions and two-dimensional linear matrices are isomorphic. We argue that the split-quaternion representation allows for easier system classification and computation of the system solution. In addition, we show that the shape of the integral curves of an underdamped system can be easily extracted from the associated split-quaternion, in comparison to the conventional procedure using complex eigenvectors.

First, we introduce the notion of split-quaternions in \cref{sec:general_quat} and how they correspond to matrices. Second, in \cref{sec:system_classification}, we relate the split-quaternions to general two-dimensional linear dynamical systems. Finally, in \cref{sec:mechanical}, we use the damped harmonic oscillator as a representative system so as to give physical interpretation to the split-quaternion representation and analyze the shape of the integral curves. 

The discussion in \cref{sec:general_quat} mainly concerns facts about split-quaternions that have been established in the past. Their application to dynamical systems (i.e. the subject of \cref{sec:system_classification} and \cref{sec:mechanical}) has not been researched previously; as such, the results presented in these sections are all new.

%% GENERAL
\input{main/quaternions/general}

%% MECHANICAL
\input{main/quaternions/mechanical}

%% NOMENCLATURE
