\chapter{Geometric Structures in Dissipative Mechanics}
\label{chap:geometric_structures}

In this chapter, we develop a geometric structure that can be used a construct Hamiltonian formalism for mechanical systems with dissipative elements. This formalism may also be applied to economic engineering systems using the analogies outlined in \cref{chap:symplectic_economics}.

Throughout this chapter, we progressively extend the conventional Hamiltonian formalism to contact and Jacobi manifolds.

First, in \cref{sec:symplectic}, the conventional Hamiltonian theory based on symplectic manifolds is explained. This section serves as a technical introduction to the rest of the chapter and does not contain new contributions.

Second, in \cref{sec:contact}, we look at contact Hamiltonian systems defined on contact manifolds. These have been proposed by some authors for dissipative mechanical systems (in particular, the damped harmonic oscillator) in the past, see e.g. \citet{Bravetti2015}. However, we propose a different formulation of the contact structure based on physical reasoning, rather than a purely mathematical argument. In addition, we show how the widely adopted Caldirola-Kanai model emerges from the symplectification of the proposed contact Hamiltonian system.

Third, in \cref{sec:jacobi}, we point out by means of an example why contact Hamiltonian systems are not able to describe general (multi-degree of freedom) mechanical systems. A (slight) generalization of contact manifolds is required instead. These manifolds belong to the general class Jacobi manifolds, which also symplectic and contact manifolds as particular subclasses.

The extension towards Jacobi manifolds has been proposed by \citet{ciaglia2018}, but it has received little to no additional attention in literature --- at least, not regarding their application to mechanical systems. We formulate the Jacobi structure using the same physical reasoning as for the contact Hamiltonian systems to make it suitable for engineering applications.

\input{main/contact/symplectic}

\input{main/contact/contact}

\input{main/contact/jacobi}
